Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models

### 2018-05-31 paper.tex
Update on Overleaf.

% Dr. Clémentine Cottineau is a research associate at the Centre for Advanced Spatial Analysis of University College London. She studied geography and economics at University Paris 1 Panthéon-Sorbonne and holds a PhD in geography from this university. Her thesis aimed at modelling urbanisation in the post-Soviet space with modular agent-based models. Since 2014, she has been working at the Centre for Advanced Spatial Analysis on urban economics and parametric scaling in the context of European systems of cities, using tools from complexity science to investigate inequality in cities and economic disparities between cities.Conc

@@

%My coauthors and I are pleased to submit an original research article entitled ``Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models'' for consideration for publication in Environment and Planning B.

%This article deals with the central yet poorly explored issue of the sensitivity of geosimulation models to their spatial context (i.e. the spatial configuration of agents and their environment), and more precisely their spatial initial conditions. We think the subject in general and the novelty of the paper in particular will meet the interest of the readers of EPB, a journal dealing with urban modelling and city science.

%Despite some preliminary results being presented as at ECTQG 2015, Bari, and at Geocomputation Conference 2017, Leeds, we confirm that neither the manuscript nor any parts of its content are currently under consideration or published in another journal. We did not have any previous interaction with EPB regarding this manuscript. We have no conflicts of interest to disclose. We do not have any opposed reviewer.

@@

\begin{abstract}
%\noindent
%\setlength{\parindent}{0pt}
Although simulation models of geographical systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain until their results are proven robust and representative of 'real-world' conditions. Sensitivity analysis usually includes the analysis of the effect of stochasticity on the variability of results, as well as the effects of small parameter changes. However, initial spatial conditions are usually taken for granted in geographical models, thus leaving completely unexplored the effect of spatial arrangements on the interactions of agents and of that of agents with their environment. In this paper, we present a method to assess the effect of initial spatial conditions on simulation models, using a systematic generator controlled by meta-parameters in order to create the density grids with which spatial simulation models are initialised. We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of resource harvesting) and a straightforward open-source work-flow using high performance computing, that the effect of space is significant in the two types of simulation. Furthermore, this effect is sometimes larger than the effect of parameters' value change. 
\end{abstract}

@@

@@

Simulation within computer models has been recognised and increasingly used by geographers as an efficient tool to explore geographical processes, hypotheses and predictive scenarios within virtual laboratories \citep{batty1971modelling, batty2007model, carley1999generating, Quesneletal2009}. %\comment{+ Seb \& James Doran in the 1970s??}
 Simulation also appears as a way to overcome the difficult analytic resolution of many spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of social and ecological systems in space and time. The specificity of geographical models compared to other social sciences' models is that space and spatial interactions are given a prime role, as geographers are driven by an explicit interest in studying the way space influences the outcomes of the model. Indeed, they usually seek to understand and model the role of space in social interactions and environmental processes, in a city for example. We think simulation can become a very good tool to achieve this, provided that models do include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of output variations to the way space is modelled. Unfortunately, the first condition is not always met, and the second is seldom even mentioned. This paper aims to fill a methodological and conceptual gap, which is a systematic testing of the sensitivity of a model's outcomes to its initial spatial conditions. To demonstrate the genericity of our approach, we develop two applications with classic simulation models commonly used as case studies for comparing and aligning simulation models \citep{Axtelletal1996, wilensky2007making}: \citet{schelling1971dynamic}'s model of segregation and \citet{EpsteinAxtell1996}'s Sugarscape model.\\

@@

An urban system can be defined as a large number of diverse social agents interacting with one another, within a limited portion of space. Social agents thus constitute the microscopic level of urban systems, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects. Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the vocabulary of complex systems, cities also exhibit emergent properties at macroscopic scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009}, which can be simulated through microscopic interactions between agents~\citep{Wu2002, Batty2007}. Complexity is partially due to bifurcations, which are determinant in spatial systems~\citep{Wilson1981, Wilson2002}. Indeed, in spatially explicit simulation models, the non-linearity of local interactions is very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interpret the resulting global structures. \\

Another crucial aspect of spatio-temporal complex systems in general, and cities in particular, is their non-ergodicity~\citep{pumain2012urban}, which means that cross-sectional samples in space (the different cities of a country at time t) are not equivalent to samples in time (the same city at time t, t+1, t+2...) to compute statistics such as average. Similarly to what \citet{gell1995quark} calls \emph{frozen accidents} in complex system generally, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system. Therefore, strong spatio-temporal path-dependencies in the trajectory of individual cities and changing social environments over time prohibits the use of ergodic models... which are the main models to be found in geosimulation! \\
% détail de chez détail : j'enlèverais le ! les ... parlent d'eux-mêmes il me semble.

Finally, although this may seem obvious, cities are not regular grids, and the distribution of density (of jobs, residents, buildings, etc.) in cities is far from isotropic, even in the most sprawled cities. On the contrary, there is a significant diversity in the way people, activities and structures are distributed in cities. In Europe for example, \citet{LeNechet2015} quantifies and classifies six broad types of residential density distributions. However, most geographical models, especially cellular automata, represent cities as uniform grids of isotropic densities. Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents is still approximated by a constant density \citep{arribas2014diverse}, although previous research shows that it is computationally and methodologically feasible to use accurate locations in a simple model such as Schelling's \citep{benenson2002entity}. The isotropic simplification is potentially harmful to the representation of the urban processes under inquiry because density and accessibility have environmental, economic and social consequences. Additionally, we expect the initial spatial distribution of agents to influence simulation results in the long run \citep{Castellanoetal2009}, because the agents' rule of action itself may depend on the spatial structure of the environment. For example, households can have different preferences with respect to the built-environment they want to live in \citep{SpielmanHarrison2014}, or agents can move around and sense their environment up to a certain distance, so that the resulting number of sensed objects will depend on the topology \citep{Banos2012} and distribution of density of the sensed environment \citep{LauriJaggi2003, FossettDietrich2009}. The way modellers abstract space is therefore a central element of geographical simulation models. A meaningful way of abstracting space might thus be to consider, not necessarily the peculiarities of every city, but at least their broad density structures.

@@

In this paper, we suggest a way of performing a sensitivity analysis to the spatial conditions of models by using a spatial generator to produce a variety of density grids, which are then fed into simulation models at initialisation. The generator being controlled by meta-parameters, we can then relate the parameters used to generate initial spatial conditions to the variation in outcome of the simulations. The purpose is two-fold: 1/ to test the robustness of simulation results to small variations of meta-parameters and 2/ to study the non-trivial effects of typical categories of spatial distribution (monocentric \textit{vs.} polycentric for example) on the results of a given model. Our approach allows for a systematic comparison of several aspects of the spatial configuration problem, which have been suggested by \citet{filatova2013spatial}, but hardly implemented and achieved in previous studies. In particular, it is applied to the effects of urban form on simulation results, using Schelling's model as a first case study and Sugarscape as a second one.

@@

Considerations of such issues in the geographical modelling literature are rather scarce, although there have been some noticeable attempts to evaluate the effects of the same set of mechanisms run from different initial spatial conditions. Mainly, they have focused on the impact of:
\begin{itemize}
\item The accuracy of geo-localised input data;
\item The shape, precision and boundaries of the modelled spatial system;
\item The degree of spatial heterogeneity modelled.
\end{itemize}
We detail these approaches in the paragraphs below. Other approaches, not included in the scope of this article, include the effects of the topology of space on the processes (cf. \citet{moreno2009integrating} who explore the role of the topology of possible interactions in a Schelling model). Generally, sensitivity analyses focus either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.); either on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time.

\paragraph{Geo-localised input data accuracy:} \citet{Thomasetal2017} show that data selection in LUTI model is inter-related with the delineation of the spatial system boundaries and scale of analysis. They provide a few examples of how the use of Exploratory Spatial Data Analysis (ESDA) prior to simulation runs can help avoiding measurement errors of model behaviour and outcomes. In the context of spatial interaction models, \citet{hagen2012new} acknowledge the dilemma between spatial resolution and computational burden, and suggest a method of adaptive zoning (where the size of destination zones depends on the distance to origin) to solve it.

\paragraph{Spatial system shape, precision, and boundaries:} \citet{Axtelletal1996} highlight the sensitivity of the average number of stable cultural regions generated to the effect of the territory width implemented in a version of the Sugarscape model which is docked (i.e. made equivalent to) to Axelrod Culture Model. \citet{FlacheHegselmann2001} show that chances for random emergence of a stable cluster of similar agents in a Schelling-like model are higher in a rectangular grid and lower in a hexagonal grid and that an irregular (Voronoi-diagram) city lattice structure favours migration stabilisation around decentralised clusters of similar agents. \citet{Banos2012} compares the behaviour of Schelling segregation model on city lattices formalized as either grid, random, scale-free and Sierpinski networks and concludes that the presence of cliques in graph-based urban structures favor segregationist behaviours. \citet{LeTexierCaruso2017}, using a set of different theoretical spatial systems, demonstrate the impact of the regularity and aggregation levels, or centrality/periphery effects, on spatial diffusion dynamics of euro coins. Similar issues were also dealt with in physical sciences, as for example \cite{horritt2001effects} studies the effects of grid cell size on the behaviour of a raster flood model and shows that increasing resolution does not increase model prediction performance under a certain limit. Similar conclusions are obtained by \cite{vazquez2002effect}, unveiling an intermediate optimal spatial resolution regarding model performance and computation time. Spatial resolution also plays a role in the Schelling model: \citet{Singhetal2009} show that the segregation patterns for certain tolerance values are strictly a small city phenomenon (8x8 city-lattice) and do not work for a larger spatial lattice (100x100), where segregation appears only for certain combinaisons of tolerance threshold and vacancy density values.

@@

\paragraph{Spatial heterogeneity:} \citet{StaufferSolomon2007} introduce asymetric interactions and empty residences in Schelling's model run on a large and regular lattice. They reveal conjoint and non-linear effects on the vacancy rates and tolerance levels on segregation patterns. \citet{Gauvinetal2010} run Schelling's segregation process in an open city-lattice to study how the variations in tolerance levels, vacancy rates and city attractiveness may create lines of vacancy lots between clusters of agents. They conclude on the functional role of vacancies, which allow weakly tolerant agents to live and be satisfied in a city environment they nevertheless perceive as hostile. \citet{HatnaBenenson2012} show that their model replications run on a 50x50 torus with 2\% of empty cells were not sensitive to the initial patterns (random and fully segregated distribution of agents). In ecology, \citet{smith2002predicting} study the spread of a disturbance in an heterogeneous landscape using a percolation model, and show that landscape structure has a significant influence on final patterns of contamination outcome. In a spatial epidemics model for an infectious disease, parametrised on real data, \citet{smith2002predicting} finds that the physical landscape heterogeneity, in particular the presence of rivers, locally influence the propagation speed.

%\paragraph{Combination:}

%\cite{ligmann2013spatially}

In this paper, we build on these contributions and go further by systematically measuring the impact of the city density distribution on model behaviour, which has so far only been assessed through a small set of illustrative spatial structures. We illustrate the potential generalisation of our approach by running two distinctive agent-based models: Schelling's model and Sugarscape.

@@

The general method workflow of our method is illustrated in Figure \ref{fig:method}. In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with various values of its parameters, we introduced a spatial generator which depends on its own meta-parameters and feeds the model $\mu$ at initialisation (lower branch). The resulting configurations can be clustered into qualitative types of spatial patterns, so that the sensitivity analysis relates to the model's parameters, how the density spatial distribution was generated and the pattern of density in order to produce qualitative conclusion about the model and its behaviour. In particular, we want to emphasize that spatial effects only derive not only from grid size or shape effects, but also from the heterogeneity of relevant geographic entities (people, housing, networks, etc.) in geographical space. In the models we produced, the initial spatial configurations  can be either flat or heterogeneous, monocentric or polycentric, based on external databases and on internal modelling (generation of synthetic population for instance \citep{bhat1999activity}).
%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[htbp] \begin{center} 
\resizebox{0.9\textwidth}{!}{ 
	\includegraphics{figures/SchemaMeta_1.png}
} \caption{\textbf{General workflow of our method.}} \label{fig:method} \end{center} \end{figure*} %
%%%%%%%%%%%%%%%%%%%%%%

In order to test for the influence of initial spatial conditions on model outputs, we use a systematic method to compare \emph{phase diagrams}. We define a phase diagram as the vector of model outputs considered as a function of model parameters. Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values $\vec{P}$ as the median of the final values of an output indicator $O$ obtained for the replications of the model initialized with $\vec{P}$.\\

%Comparing phase diagrams is useful because it allows us to compare the 
We have as many phase diagrams than we have spatial grids, which makes a qualitative visual comparison not realistic (with around 50 different spatial configurations for each model experiment). A solution is to use systematic quantitative procedures to compare them. To our knowledge there exist no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature. We introduce a measure of the relative distance $d_r$ between two phase diagrams $\mu_{\vec{\alpha}_1}$ and $\mu_{\vec{\alpha}_2}$. Phase diagrams are denoted by a common function $\mu$ indexed by what we call \emph{meta-parameters} $\vec{\alpha}$, that capture the spatial configuration (in practice these can be parameters of an upstream model to generate the configuration, or a description of the configuration itself). The $d_r$ measure is taken as the share of between-diagrams variability relative to their internal variability, given formally in the case of a one-dimensional phase diagram by

@@

where $\mu_{\vec{\alpha}_i}$ are the phase diagrams at given meta-parameters $\vec{\alpha}_i$. $d$ is a functional distance that we take simply as Euclidian distance. The variances are estimated within each phase diagram $f_{\vec{\alpha_i}}$. For a multi-dimensional phase diagram, we sum these relative distances over the components.

The last methodological point which we need to emphasize is the relation between the present workflow and model exploration workflows in general. The ideas of multi-modeling and extensive model exploration are nothing from new as ~\cite{openshaw1983data} already advocated for ``model-crunching'', but their effective use only begins to emerge thanks to the development of new methods and tools together with an explosion of computation capabilities. The model exploration platform OpenMOLE~\citep{reuillon2013openmole} allows to embed any model as a blackbox, to write modulable exploration workflow using advanced methodologies such as genetic algorithms and to distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids. In our case, this tool is a powerful way to embed both the sensitivity analysis and the meta-sensitivity analysis, and to allow the coupling of any spatial generator with any model in a straightforward way as long as the model can take its spatial configuration as an input or from an input file. In this paper, we use the OpenMOLE platform for the spatial environment and the model coupling, placing ourselves in the renewed framework of multi-modelling \citep{cottineau2015modular}.
%%%%%%%%%%%%%%%%%%%%%%
\subsection{Spatial generator of density grids}

>>>>

% Dr. Clémentine Cottineau is a research associate at the Centre for Advanced Spatial Analysis of University College London. She studied geography and economics at University Paris 1 Panthéon-Sorbonne and holds a PhD in geography from this university. Her thesis aimed at modelling urbanisation in the post-Soviet space with modular agent-based models. Since 2014, she has been working at the Centre for Advanced Spatial Analysis on urban economics and parametric scaling in the context of European systems of cities, using tools from complexity science to investigate inequality in cities and economic disparities between cities.

@@

%My coauthors and I are pleased to submit an original research article entitled ``Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models'' for consideration for publication in Computer, Environment and Urban Systems.

%This article deals with the central yet poorly explored issue of the sensitivity of geosimulation models to their spatial context (i.e. the spatial configuration of agents and their environment), and more precisely their spatial initial conditions. We think the subject in general and the novelty of the paper in particular will meet the interest of the readers of CEUS, a journal dealing with computerised models and urban systems.

%Despite some preliminary results being presented as at ECTQG 2015, Bari, and at Geocomputation Conference 2017, Leeds, we confirm that neither the manuscript nor any parts of its content are currently under consideration or published in another journal. We did not have any previous interaction with CEUS regarding this manuscript. We have no conflicts of interest to disclose. We do not have any opposed reviewer.

@@

\begin{abstract}
%\noindent
%\setlength{\parindent}{0pt}
Although simulation models of geographical systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain \replaced{unless}{until} their results are proven robust and representative of 'real-world' conditions. Sensitivity analysis usually includes the analysis of the effect of stochasticity on the variability of results, as well as the effects of small parameter changes. However, initial spatial conditions are usually taken for granted in geographical models, thus leaving completely unexplored the effect of spatial arrangements on the interactions of agents and of that of agents with their environment. In this paper, we present a method to assess the effect of \added{some} initial spatial conditions on simulation models, using a systematic generator controlled by meta-parameters in order to create \deleted{the} density grids with which spatial simulation models are initialised. We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of \replaced{unequal societies}{resource harvesting}) and a straightforward open-source work-flow using high performance computing, that the effect of \replaced{initial spatial arrangements}{space} is significant in the two \replaced{models}{types of simulation}. Furthermore, this effect is sometimes larger than the effect of parameters' value change. 
\end{abstract}

@@

\todo{Response to reviewers
\begin{itemize}
\item For reviewer 1, explicitly mention the simulation framework at the beginning (and position it regarding others such as dynamical systems) with literature
\item Similarly, clarify the aggregation step, how the indicators are macro, have a thematic meaning, and distribution converge clearly (possibly : add Sharpe ratios statistics and type of distrib for better fits)
\item Rev 1 : (complementary to above) : positioning regarding some analytical works / cite some ? Precise that we do not aim at understanding fully model behavior, but finding some sensitivity to space
\item Rev 2 : precise which Schelling we use.
\item On real configs that can not be modified : remark by Marion on MAUP etc ; and add in discussion a layus on synthetic data resembling real configs etc.
\item Remove or adapt the concept of validation to avoid the epistemological trap
\item To be debated : do we keep the detailed description of the spatial generator ? (knowing that it is otherwise fully described and explored here : \url{https://arxiv.org/abs/1708.06743})
\end{itemize}
}

\textbf{Example track changes : }\added{this is added text}, \deleted{removed} text or \replaced{new}{old} replaced text

@@

\added{Computer} simulation \deleted{within computer models} has been recognised and increasingly used by geographers as an efficient tool to explore geographical processes, hypotheses and predictive scenarios within virtual laboratories \citep{batty1971modelling, batty2007model, carley1999generating, Quesneletal2009}. %\comment{+ Seb \& James Doran in the 1970s??}
 Simulation also appears as a way to overcome the difficult analytic resolution of many spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of \added{path-dependent} social and ecological systems\deleted{ in space and time}. The specificity of geographical models compared to other social sciences' models is that space and spatial interactions are given a prime role, as geographers are driven by an explicit interest in studying the way space influences the outcomes of the model \replaced{in}{. Indeed, they usually seek to} understand\replaced{ing}{ and model} the role of space \replaced{on}{in} social interactions and environmental processes, in a city for example. We think simulation can become a very good tool to achieve this, provided that models \deleted{do} include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of output variations to the way space is \replaced{represented}{modelled}. Unfortunately, the first condition is not always met, and the second is seldom even mentioned. This paper aims to fill a methodological and conceptual gap, which is a systematic testing of the sensitivity of a model's outcomes to its initial spatial conditions. To demonstrate the genericity of our approach, we develop two applications with classic simulation models commonly used as case studies for comparing and aligning simulation models \citep{Axtelletal1996, wilensky2007making}: \citet{schelling1971dynamic}'s model of segregation and \citet{EpsteinAxtell1996}'s Sugarscape model.\\

@@

\replaced{Geographical systems}{An urban system} can be \replaced{crudely described as}{defined as a large number of diverse} social agents interacting with one another, within a limited portion of space. Social agents thus constitute the microscopic level of \replaced{the}{urban} system\deleted{s}, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects\added{, also known as path-dependent \citep{arthur1994increasing}. Therefore, observing a system at different points in time does not equate to observing different system at a single point in time. This general property of ergodicity applies to geographical elements such as road networks or built-up areas \citep{pumain2003approche}}. Similarly to what \citet{gell1995quark} calls \emph{frozen accidents} in complex system generally, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system. Therefore, strong spatio-temporal path-dependencies in the trajectory of individual cities and changing social environments over time prohibits the use of ergodic models... which are the main models to be found in geosimulation.

Self-organization has been shown to be a central feature of \replaced{geographical}{urban} systems \added{in general and cities in particular} ~\citep{AllenSanglier1981,saint1989villes, Portugali2000}. In the vocabulary of complex systems, cities also exhibit emergent properties at macroscopic scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009}, which can be simulated through microscopic interactions between agents~\citep{Wu2002, Batty2007}. Complexity is partially due to bifurcations, which are determinant in spatial systems~\citep{Wilson1981, Wilson2002}. Indeed, in spatially explicit simulation models, the non-linearity of local interactions is very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interpret the resulting global structures. \added{In that sense, the impact of initial \emph{spatial} settings on final outcomes is assumed to be significant just as other initial conditions, but of more interest to the geographer}. \\

\deleted{Another crucial aspect of spatio-temporal complex systems in general, and cities in particular, is their non-ergodicity, which means that cross-sectional samples in space (the different cities of a country at time t) are not equivalent to samples in time (the same city at time t, t+1, t+2...) to compute statistics such as average.}

Finally, although this may seem obvious, cities are not regular grids, and the distribution of density (of jobs, residents, buildings, etc.) \deleted{in cities} is far from isotropic, even in the most sprawled cities. On the contrary, there is a significant diversity in the way people, activities and structures are distributed in cities. In Europe for example, \citet{LeNechet2015} quantifies and classifies six broad types of residential density distributions. However, most geographical models, especially cellular automata, \added{still} represent cities as uniform grids of isotropic densities. Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents is still approximated by a constant density \citep{arribas2014diverse}, although previous research shows that it is computationally and methodologically feasible to use accurate locations in a simple model such as Schelling's \citep{benenson2002entity}. The isotropic simplification is potentially harmful to the representation of the urban processes under inquiry because density and accessibility have environmental, economic and social consequences. Additionally, we expect the initial spatial distribution of agents to influence simulation results in the long run \citep{Castellanoetal2009}, because the agents' rule of action itself may depend on the spatial structure of the environment. For example, households can have different preferences with respect to the built-environment they want to live in \citep{SpielmanHarrison2014}, or agents can move around and sense their environment up to a certain distance, so that the resulting number of sensed objects will depend on the topology \citep{Banos2012} and distribution of density of the sensed environment \citep{LauriJaggi2003, FossettDietrich2009}. The way modellers \replaced{represent}{abstract} space is therefore a central element of geographical simulation models. \added{However, this step is rarely explicit and authors tend to generalise the conclusion of the model from a single spatial case study (be it a theoretical pattern or a singular empirical one}. A meaningful way of \replaced{representing}{abstracting} space might thus be to consider, not necessarily the peculiarities of every city, but at least their broad density structures\added{ so as to estimate the variability of the model behaviour in plausible spatial arrangements}.

@@

In this paper, \replaced{we offer a methodological solution to the problem of sensitivity of simulation models to initial spatial conditions with two application cases (Schelling and Sugarscape). In no way do we want to provide a full exploration of these two particular models, their attractors and/or potential policy implications. Instead, we present}{we suggest} a way of performing a sensitivity analysis to the spatial conditions of models by using a spatial generator to produce a variety of density grids, which are then fed into simulation models at initialisation. The generator being controlled by meta-parameters, we can then relate the parameters used to generate initial spatial conditions to the variation in outcome of the simulations. The purpose is two-fold: 1/ to test the robustness of simulation results to small variations of meta-parameters and 2/ to study the non-trivial effects of typical categories of spatial distribution (monocentric \textit{vs.} polycentric for example) on the results of a given model. Our approach allows for a systematic comparison of several aspects of the spatial configuration problem, which have been suggested by \citet{filatova2013spatial}, but hardly implemented and achieved in previous studies. In particular, it is applied to the effects of urban form on simulation results, using Schelling's model as a first case study and Sugarscape as a second one.

@@

Considerations of such issues in the geographical modelling literature are rather scarce.\replaced{Despite the opportunity of simulation in the representation of space and its potential observation biases,}{ although} there have been some noticeable attempts to evaluate the effects of the same set of mechanisms run from different initial spatial conditions\replaced{, which}{. Mainly, they} have focused on the impact of:
\begin{itemize}
\item The accuracy of geo-localised input data;
\item The shape, precision and boundaries of the modelled spatial system;
\item The degree of spatial heterogeneity modelled.
\end{itemize}

\deleted{We detail these approaches in the paragraphs below. Other approaches, not included in the scope of this article, include the effects of the topology of space on the processes (cf. moreno who explore the role of the topology of possible interactions in a Schelling model). }

\paragraph{Geo-localised input data accuracy} \citet{Thomasetal2017} show that data selection in LUTI model is inter-related with the delineation of the spatial system boundaries and scale of analysis. They provide a few examples of how the use of Exploratory Spatial Data Analysis (ESDA) prior to simulation runs can help avoiding measurement errors of model behaviour and outcomes. In the context of spatial interaction models, \citet{hagen2012new} acknowledge the dilemma between spatial resolution and \added{the} computational burden, and suggest a method of adaptive zoning (where the size of destination zones depends on the distance to origin) to solve it.

\paragraph{Spatial system shape, precision, and boundaries} \citet{Axtelletal1996} highlight the sensitivity of the average number of stable cultural regions generated to the effect of the territory width implemented in a version of the Sugarscape model which is docked (i.e. made equivalent to) to Axelrod Culture Model. \citet{FlacheHegselmann2001} show that chances for random emergence of a stable cluster of similar agents in a Schelling-like model are higher in a rectangular grid and lower in a hexagonal grid and that an irregular (Voronoi-diagram) city lattice structure favours migration stabilisation around decentralised clusters of similar agents. \citet{Banos2012} compares the behaviour of Schelling segregation model on city lattices formalized as either grid, random, scale-free and Sierpinski networks and concludes that the presence of cliques in graph-based urban structures favours segregationist behaviours. \citet{LeTexierCaruso2017}, using a set of different theoretical spatial systems, demonstrate the impact of the regularity and aggregation levels, or centrality/periphery effects, on spatial diffusion dynamics of euro coins. Similar issues were also dealt with in physical sciences, as for example \cite{horritt2001effects} studies the effects of grid cell size on the behaviour of a raster flood model and shows that increasing resolution does not increase model prediction performance under a certain limit. Similar conclusions are obtained by \cite{vazquez2002effect}, unveiling an intermediate optimal spatial resolution regarding model performance and computation time. Spatial resolution also plays a role in the Schelling model: \citet{Singhetal2009} show that the segregation patterns for certain tolerance values are strictly a small city phenomenon (8x8 city-lattice) and do not work for a larger spatial lattice (100x100), where segregation appears only for certain combinaisons of tolerance threshold and vacancy density values.

@@

\paragraph{Spatial heterogeneity} \citet{StaufferSolomon2007} introduce asymetric interactions and empty residences in Schelling's model run on a large and regular lattice. They reveal conjoint and non-linear effects on the vacancy rates and tolerance levels on segregation patterns. \citet{Gauvinetal2010} run Schelling's segregation process in an open city-lattice to study how the variations in tolerance levels, vacancy rates and city attractiveness may create lines of vacancy lots between clusters of agents. They conclude on the functional role of vacancies, which allow weakly tolerant agents to live and be satisfied in a city environment they nevertheless perceive as hostile. \citet{HatnaBenenson2012} show that their model replications run on a 50x50 torus with 2\% of empty cells were not sensitive to the initial patterns (random and fully segregated distribution of agents). In ecology, \citet{smith2002predicting} study the spread of a disturbance in an heterogeneous landscape using a percolation model, and show that landscape structure has a significant influence on final patterns of contamination outcome. In a spatial epidemics model for an infectious disease, parametrised on real data, \citet{smith2002predicting} finds that the physical landscape heterogeneity, in particular the presence of rivers, locally influence the propagation speed.

%\paragraph{Combination:} 
 Generally, sensitivity analyses \replaced{have focused}{focus} either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.); either on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time.
%\cite{ligmann2013spatially}

In this paper, we build on these contributions and go further by systematically measuring the impact of the city density distribution \replaced{aggregate model behaviours (e.g. the final segregation level for Schelling)}{on model behaviour, which has so far only been assessed through a small set of illustrative spatial structures}. We illustrate the potential generalisation of our approach by running two distinctive agent-based models: Schelling's model and Sugarscape.

@@

The general method workflow of our method is illustrated in Figure \ref{fig:method}. In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with \replaced{different}{various} values of its parameters, we introduce a spatial generator which depends \deleted{on its} own meta-parameters and feeds the model $\mu$ at initialisation (lower branch). The resulting configurations \emph{can be} clustered into qualitative types of spatial patterns.\deleted{so that} The sensitivity analysis relates to the model's parameters \added{to} how the density spatial distribution was generated and the pattern of density \replaced{generated in order to produce qualitative knowledge from}{in order to produce qualitative conclusion about} the model and its behaviour. In particular, we want to emphasize that spatial effects \deleted{only} derive not only from grid size or shape effects, but also from the heterogeneity of \replaced{the distribution of}{relevant} geographical entities (people, housing, networks, etc.) in \deleted{geographical} space. In the models we \replaced{used as examples}{produced}, the initial spatial configurations  can be either flat or heterogeneous, monocentric or polycentric, based on external databases and on internal modelling (generation of synthetic population for instance \citep{bhat1999activity}).
%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[htbp] 
\begin{center} 
\resizebox{0.9\textwidth}{!}{ 
	\includegraphics{figures/SchemaMeta_1.png}
} \caption{\textbf{General workflow.}} \label{fig:method}
\end{center}
\end{figure*} %
%%%%%%%%%%%%%%%%%%%%%%

In order to test the influence of initial spatial conditions on model outputs, we use a systematic method to compare \emph{phase diagrams}. \added{Following} \citet{Gauvinetal2009}, we define a phase diagram as the vector of \added{final aggregated} model outputs considered as a function of model parameters. We have as many phase diagrams than we have spatial grids, which makes a qualitative visual comparison not realistic (with around 50 different spatial configurations for each model experiment). A solution is to use systematic quantitative procedures to compare them.
Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values $\vec{P}$ as the median of the final values of an output indicator $O$ obtained for the replications of the model initialized with $\vec{P}$.\\

To our knowledge there exist no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature. We introduce a measure of the relative distance $d_r$ between two phase diagrams $\mu_{\vec{\alpha}_1}$ and $\mu_{\vec{\alpha}_2}$. Phase diagrams are denoted by a common function $\mu$ indexed by what we call \emph{meta-parameters} $\vec{\alpha}$, that capture the spatial configuration \footnote{in practice these can be parameters of an upstream model to generate the configuration, or a description of the configuration itself}. The $d_r$ measure is taken as the share of between-diagrams variability relative to their internal variability, given formally in the case of a one-dimensional phase diagram by

@@

where $\mu_{\vec{\alpha}_i}$ are the phase diagrams at given meta-parameters $\vec{\alpha}_i$. $d$ is a functional distance that we take simply as Euclidian distance. The \replaced{internal variabilities}{variance} are estimated \added{as the variance} within each phase diagram $f_{\vec{\alpha_i}}$. For a multi-dimensional phase diagram, we sum these relative distances over the components.

The last methodological point which we need to emphasize is the relation between the present workflow and model exploration workflows in general. The ideas of multi-modeling and extensive model exploration are nothing from new as ~\cite{openshaw1983data} already advocated for ``model-crunching'', but their effective use only begins to emerge thanks to the development of new methods and tools together with an explosion of computation capabilities. The model exploration platform OpenMOLE~\citep{reuillon2013openmole} allows to embed any model as a blackbox, to write modulable exploration workflow using advanced methodologies such as genetic algorithms and to distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids. In our case, this tool is a powerful way to embed both the sensitivity analysis and the meta-sensitivity analysis, and to allow the coupling of any spatial generator with any model in a straightforward way as long as the model can take its spatial configuration as an input or from an input file. In this paper, we use the OpenMOLE platform for the spatial environment and the model coupling, placing ourselves in the  framework of multi-modelling \citep{cottineau2015modular}.

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Spatial generator of density grids}
